Modular analysis of gene expression data with R

### 2009-12-11 isanote/modular.tex
Updated isa application note.

In this short note, we introduce two BioConductor \citep{BioC}, software
packages written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
implementation of the ISA and~\Rpackage{eisa} provides a convenient interface
to run the ISA and visualize its output.

\section{Availability:}
The ISA homepage is at \href{http://www.unil.ch/cbg/homepage/software.html}%
{\url{http://www.unil.ch/cbg/homepage/software.html}}
\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
\end{abstract}

@@

The ISA identifies modules by an iterative procedure. The algorithm starts
from an input seed (corresponding to some set of genes or samples) which is
refined at each iteration by adding and/or removing genes and/or samples
until the process converges to a stable set, which is referred to as a
transcription module. This iterative procedure is typically performed to a
large number of seeds. In the unsupervised approach these seeds are chosen
randomly to sample uniformly the immense search space. We also implemented a
semi-supervised method, to which we refer as ``smart seeding'', where the
seeds a biased to start with certain sets of genes or samples for which
there is prior knowledge.

The output of ISA is a collection of potentially overlapping modules. Each
gene and each sample is attributed a score that reflects the strength of the
association with the module.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{\subfig{(A)} Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package. The input of the \Rpackage{eisa} pipeline is a
  BioConductor \Rclass{ExpressionSet} object. This is normalized to
  $Z$-scores (1); the ISA iterations are performed (2); and the
  biclusters are then merged and filtered based on the robustness
  measure (3). The \Rpackage{eisa} package contains tools to perform
  many GO, KEGG, etc. enrichment calculations quickly (4), and several
  bicluster visualization tools (5).
  Subfigures \subfig{B} and \subfig{C} were generated
  using the acute lymphoblastic leukemia data set, see
  \citep{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(B)} Heatmap for a single module, showing correlated
  expression for the genes and samples. The red lines are the gene and
  sample scores.
  \subfig{(C)} A module tree. Each rectangle is a bicluster; see the
  definition of the edges in the text. The modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the bicluster. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
}
\label{fig:workflow}
\end{figure*}

@@

% Batch correction
\emph{Batch correction.}
To study the global organization of a transcription program including
many aspects of transcriptional regulation one often combines several
microarray experiments into a single dataset. In such a case it is important
to properly normalize such data and reduce the bias due to constituent
datasets. Several methods address this challenge,
see e.g.~\citep{johnson07} for an algorithm that has a GNU R
implementation. 
% (S: Do we have a standard flag to call their software?)

@@

% Normalization.
\emph{ISA normalization.}
In each iteration of the ISA weighted sums of expression levels over either
genes or samples are computed. Since different genes
typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
calculated for each gene across all samples, the other for each sample
across all genes.

% Random and smart seeding.
\emph{Random and smart seeding.}
As mentioned previously, ISA can be performed with random or smart
seeds, based on the application.

% The ISA iteration.
\emph{The ISA iteration.}
This step is the core of the ISA and finds biclusters from the
starting seeds.

% Merging the modules.
\emph{Merging the modules.}
It is possible that several seeds converge to the same, or very similar
biclusters. This step eliminates such duplicates.

% The robustness measure and filtering the modules.
\emph{Robustness of biclusters.}
To access the significance of a bicluster, we designed a robustness
measure, that can be used to filter out spurious modules. This is done
based on performing ISA on the scrambled input data.

>>>>

In this short note, we introduce two BioConductor \citep{BioC}, software
packages written in GNU~R: The~\Rpackage{isa2} includes a highly optimized
implementation of the ISA and~\Rpackage{eisa} provides a convenient interface
to run the ISA, visualize its output and puts the biclusters into
biological context.

\section{Availability:}
\href{http://www.unil.ch/cbg/ISA}%
{\url{http://www.unil.ch/cbg/ISA}}
\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
\end{abstract}

@@

The ISA identifies modules by an iterative procedure. The algorithm starts
from an input seed (corresponding to some set of genes or samples) which is
refined at each iteration by adding and/or removing genes and/or samples
until the process converges to a stable set, which is referred to as a
transcription module.

The output of ISA is a collection of potentially overlapping
modules. In every module, each gene and each sample is attributed a
score that reflects the strength of the association with the module.

@@

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{\subfig{(A)} Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package. See the text for details.
  Subfigures \subfig{B} and \subfig{C} were generated
  using the acute lymphoblastic leukemia data set, see
  \citep{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(B)} Heatmap for a single module, showing correlated
  expression for the genes and samples. The red lines are the gene and
  sample scores.
  \subfig{(C)} A module tree. Each rectangle is a bicluster; see the
  definition of the edges in the text. The modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the bicluster. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
}
\label{fig:workflow}
\end{figure*}

@@

% Batch correction
\emph{Batch correction.}
To study the global organization of a transcription program including
many aspects of transcriptional regulation one often combines several
microarray experiments into a single dataset. In such a case
additional data normalization is vital to reduce the bias due to
constituent datasets. Several methods address this challenge,
see e.g.~\citep{johnson07} for an algorithm that has a GNU R
implementation. 
% (S: Do we have a standard flag to call their software?)

@@

% Normalization.
\emph{ISA normalization.} (Step 1 on Fig.~\ref{fig:workflow}.)
In each iteration of the ISA thresholded weighted sums of expression
levels over either genes or samples are computed. Since different genes
typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
calculated for each gene across all samples, the other for each sample
across all genes.

% Random and smart seeding.
\emph{Random and smart seeding, ISA iteration} (Step 2.)
The ISA iteration is typically performed to a
large number of seeds. In the unsupervised approach these seeds are chosen
randomly to sample uniformly the immense search space. We also implemented a
semi-supervised method, to which we refer as ``smart seeding'', where the
seeds a biased to start with certain sets of genes or samples based on
prior knowledge.As mentioned previously, ISA can be performed with random or smart
seeds, based on the application.

% Merging the modules.
\emph{Merging the modules.} (Step 3.)
It is possible that several seeds converge to the same, or very similar
biclusters. This step eliminates such duplicates.

% The robustness measure and filtering the modules.
\emph{Robustness of biclusters.} (Step 3.)
To access the significance of a bicluster, we designed a robustness
measure, that can be used to filter out spurious modules. This is done
based on performing ISA on the scrambled input data.
